---
title: "Analysis of lncRNAs"
output:
  html_notebook:
    toc: true
    number_sections: true
    theme: readable
author: Marat Mufteev
date: "`r format(Sys.time(), 'Generated on %d %B, %Y')`"
---

Upload libraries
```{r}
library(dplyr)
library(ggplot2)
```

Current directory to generate data paths
```{r}
current_dir = getwd()
```

Upload RNA abundance and translation status data from the following works:
A) Human heart Ribo-seq. PMID:31155234. Named here as Heart
B) ESC-derived neurons Ribo-seq. PMID:27655340. Named here as TSC2
C) iPS-based neurodevelopment TRAP-seq. From this work
```{r}
path = paste(current_dir, 
             "/data/annotation/Homo_sapiens.GRCh38.cdna.all.longer_125nt.tab", 
             sep = "")
ids = read.table(path, header = F)
colnames(ids) = c("Name", "gene", "type")
filter_names = (ids %>% 
                  filter(type == "lncRNA") %>%
                  mutate(Name = gsub("[.][0-9]|[.][0-9][0-9]", "", Name)))$Name

path = paste(current_dir, 
             "/data/TSC2_Input.tbl", 
             sep = "")
input_A = read.table(path, header = T) %>%
          mutate(Name = gsub("[.][0-9]|[.][0-9][0-9]", "", Name)) %>%
          filter(Name %in% filter_names) %>%
          select(Name, contains("TPM")) %>%
          mutate(TPM.TSC2.mean = apply(X = .[,2:5], MARGIN = 1, FUN = mean, na.rm = T),
                 TPM.TSC2.sd = apply(X = .[,2:5], MARGIN = 1, FUN = sd, na.rm = T)) %>%
          select(Name, TPM.TSC2.mean, TPM.TSC2.sd)
          
path = paste(current_dir, 
             "/data/quant.lncRNAs.stats.sf", 
             sep = "")
input_B = read.table(path, header = T) %>%
          dplyr::rename(Name = transcript) %>%
          filter(Name %in% filter_names) %>%
          select(Name, contains("TPM.Input"))
          
input_join = full_join(input_B, input_A, by = "Name") %>%
             left_join(ids %>% mutate(Name = gsub("[.][0-9]|[.][0-9][0-9]", "", Name)), 
                       by = "Name") %>%
             distinct()

path = paste(current_dir, 
             "/data/RE.deseq.horizontal.transcripts.quant_normalized.stats.tab", 
             sep = "")
# If RE > 1 then store TRUE and FALSE otherwise
RE_tbl = read.table(path, header = T) %>%
         dplyr::rename(Name = gene) %>%
         left_join(ids) %>%
         filter(type %in% "lncRNA") %>%
         select(Name, gene, contains("mean")) %>%
         mutate_if(is.double, function(x) {x > 1.0}) %>%
         select(-Name) %>%
         group_by(gene) %>%
         summarise_all(function(x) {any(x, na.rm = T)})

path = paste(current_dir, 
             "/data/humanheart_expressedLncRNAs.csv", 
             sep = "")
heart = read.csv(path, header = T) %>%
        mutate(gene = gene_name)

path = paste(current_dir, 
             "/data/Human_GSE78959_ORF.noncoding.tab", 
             sep = "")
TSC_neuron = read.table(path, header = T)

TSC_tbl = data.frame(gene = unique(TSC_neuron$GeneName),
                     TSC2 = TRUE) %>%
          left_join(ids) %>%
          filter(type %in% "lncRNA") %>%
          group_by(gene) %>%
          summarise(TSC2 = dplyr::first(TSC2))
```

Estimate overlap in number of lncRNAs (lincRNAs) between datasets
```{r}
ORF_tbl = full_join(RE_tbl, TSC_tbl) %>%
          select(gene, contains("RE"), contains("TSC2")) %>%
          mutate_all(function(x) {ifelse(is.na(x), FALSE, x)})

ORF_tbl = full_join(RE_tbl, heart %>% dplyr::rename(gene = gene_name,
                                                    Heart = translated)) %>%
          select(gene, contains("RE"), contains("Heart")) %>%
          mutate_all(function(x) {ifelse(is.na(x), FALSE, x)})

lncRNA_gene = input_join %>%
              select(gene, contains("mean")) %>%
              group_by(gene) %>%
              summarise_all(sum, na.rm = T) %>%
              na.omit()

# Compare TSC2 and TRAP-seq dataset
collector_tbl = data.frame()
for (TPM in c(0.01, 0.1, 1, 5, 10, 25, 40)) {
  for (A in c("TSC2")) {
  for (B in colnames(ORF_tbl)[-c(1, 7)]) {
    lncRNA = lncRNA_gene %>%
             left_join(heart %>% 
                         dplyr::rename(gene = gene_name,
                                       Heart = FPKM_mean_rna_gene) %>%
                         distinct(gene, Heart))
    colnames(lncRNA) = gsub("TPM[.]|Input_|[.]mean", "", colnames(lncRNA))
    
    if (A != B) {
      cellA = gsub("RE_|[.]mean", "", A)
      cellB = gsub("RE_|[.]mean", "", B)
      print(paste(cellA, cellB))
      
      lncRNA = lncRNA %>% select(gene, cellA, cellB)
      genes_common = (data.frame(lncRNA %>%
                                 mutate(RES = (.[,2] > TPM) & (.[,3] > TPM)) %>%
                                 filter(RES == TRUE)))$gene
      tblA = data.frame(ORF_tbl %>% filter(gene %in% genes_common) %>% select(A))[,1]
      tblB = data.frame(ORF_tbl %>% filter(gene %in% genes_common) %>% select(B))[,1]
      
      common = sum((tblA + tblB) > 0) 
      
      
      
      count = sum((tblA + tblB) == 2) 
      collector_tbl = bind_rows(collector_tbl,
                                data.frame(overlap = round(count/common* 100, 0),
                                           N = common,
                                           Tr = count,
                                           cellA = A,
                                           cellB = B,
                                           TPM_threshold = TPM))
    }
  }
}
}

ggplot(data = collector_tbl,
       aes(x = TPM_threshold,
           y = overlap,
           color = cellB,
           group = cellB)) +
  geom_point() +
  geom_line()
```

Bar plot to visualize overlap between datasets
```{r}
Xmin = 125
step = 25

bar_df = collector_tbl %>% filter(TPM_threshold == 5)
xlabels = gsub("TE_|[.]mean", "", bar_df$cellB)
barplot(bar_df$Tr, xlim = c(-Xmin, 50), axes = FALSE, horiz = T)
barplot(-bar_df$N, add = TRUE, names.arg = xlabels, axes = FALSE, horiz = T, las = 2)
abline(h = 0)
axis(1, lwd = 0, tick = FALSE, at = seq(-Xmin,25,step),
     labels = c(rev(seq(0,Xmin,step)), seq(step,25,step)), las = 1)
mtext("Ribo-seq", 1, line = 3, at = 12.5)
mtext("Input", 1, line = 3, at = -65.5)
```

Summary table with lncRNAs
```{r}
full_tbl = lncRNA_gene %>%
           left_join(heart %>%
                     dplyr::rename(gene = gene_name,
                                   Heart_FPKM_RNA = FPKM_mean_rna_gene,
                                   Heart_translated = translated) %>%
                     distinct(gene, Heart_FPKM_RNA, Heart_translated)) %>%
           left_join(ORF_tbl) %>%
           dplyr::rename(TSC2_translated = TSC2,
                         iPS37_translated = RE_iPS37.mean,
                         NPC37_translated = RE_NPC37.mean,
                         Neu37_translated = RE_Neu37.mean,
                         NPC20_translated = RE_NPC20.mean,
                         Neu20_translated = RE_Neu20.mean)
```







