---
title: "Analysis of transcripts sequence features"
output:
  html_notebook:
    toc: true
    number_sections: true
    theme: readable
author: Marat Mufteev
date: "`r format(Sys.time(), 'Generated on %d %B, %Y')`"
---

Upload libraries
```{r}
library(dplyr)
library(ggplot2)
```

Current directory to generate data paths
```{r}
current_dir = getwd()
```

Transcript identifiers, statistics of a sequence, RE from quantile normalization method and counts
```{r}
ids_path = paste(current_dir, 
                 "/data/annotation/grch38.transcript.genes.csv", 
                 sep = "")
seq_path = paste(current_dir, 
                 "/data/transcripts_stats.csv", 
                 sep = "")
RE_path = paste(current_dir, 
                "/data/RE.deseq.horizontal.quant_normalized.stats.tab", 
                sep = "")
TPM_path = paste(current_dir, 
                 "/data/quants.clean.stats.sf", 
                 sep = "")
TPM_3UTR_path = paste(current_dir, 
                      "/data/quant.3UTR.average.sf", 
                      sep = "")
RE_3UTR_path = paste(current_dir, 
                     "/data/RE.3UTR.deseq.horizontal.quant_normalized.stats.tab", 
                     sep = "")

ids = read.csv(ids_path, header = TRUE)
seq_stats = read.csv(seq_path, header = T) %>%
            mutate(gc_percentage.cds = gc_count.cds/length.cds,
                   gc_percentage.3utr = gc_count.3utr/length.3utr,
                   gc_percentage.5utr = gc_count.5utr/length.5utr,
                   gc_percentage.tx = (gc_count.5utr + gc_count.cds + gc_count.3utr)/(length.5utr + length.cds + length.3utr))

RE.genes = read.table(RE_path, header = TRUE) 
TPM.transcripts = read.csv(TPM_path, header = TRUE)

TPM.3UTR = read.table(TPM_3UTR_path, header = TRUE)
RE.3UTR = read.table(RE_3UTR_path, header = TRUE)
```

Combine RE and sequence statistics tables
```{r}
# Thresholds for the noise levels
genes_RE_threshold = 0.3
qapa_RE_threshold = 0.5

genes_tbl = left_join(ids, seq_stats, by = "transcript") %>%
            left_join(RE.genes, by = "gene") %>%
            mutate(RE_iPS37.mean = replace(RE_iPS37.mean, 
                                           RE_iPS37.noise > genes_RE_threshold, NA),
                   RE_NPC37.mean = replace(RE_NPC37.mean, 
                                           RE_NPC37.noise > genes_RE_threshold, NA),
                   RE_NPC20.mean = replace(RE_NPC20.mean, 
                                           RE_NPC20.noise > genes_RE_threshold, NA),
                   RE_Neu37.mean = replace(RE_Neu37.mean, 
                                           RE_Neu37.noise > genes_RE_threshold, NA),
                   RE_Neu20.mean = replace(RE_Neu20.mean, 
                                           RE_Neu20.noise > genes_RE_threshold, NA))
counts_genes_tbl = left_join(TPM.transcripts, seq_stats, by = "transcript") %>%
                   left_join(ids, by = "transcript")
utrs_tbl = left_join(TPM.3UTR %>% select(Name, Length),
                     RE.3UTR %>% dplyr::rename(Name = UTR3_isoform), by = "Name") %>%
           mutate(RE_iPS37.mean = replace(RE_iPS37.mean, RE_iPS37.noise > qapa_RE_threshold, NA),
                  RE_NPC37.mean = replace(RE_NPC37.mean, RE_NPC37.noise > qapa_RE_threshold, NA),
                  RE_NPC20.mean = replace(RE_NPC20.mean, RE_NPC20.noise > qapa_RE_threshold, NA),
                  RE_Neu37.mean = replace(RE_Neu37.mean, RE_Neu37.noise > qapa_RE_threshold, NA),
                  RE_Neu20.mean = replace(RE_Neu20.mean, RE_Neu20.noise > qapa_RE_threshold, NA))

genes_stats = genes_tbl %>%
              group_by(gene) %>%
              filter(length.cds == max(length.cds)) %>%
              arrange(length.cds)
```

RE dependence on CDS length. Analysis done on a transcript level
```{r}
genes_avg = genes_stats %>%
            mutate(length = floor(length.cds / 500)) %>%
            group_by(length) %>%
            summarise(length.cds = mean(length.cds, na.rm = TRUE), 
                      RE_iPS37 = mean(RE_iPS37.mean, na.rm = TRUE),
                      RE_NPC37 = mean(RE_NPC37.mean, na.rm = TRUE),
                      RE_Neu37 = mean(RE_Neu37.mean, na.rm = TRUE),
                      RE_NPC20 = mean(RE_NPC20.mean, na.rm = TRUE),
                      RE_Neu20 = mean(RE_Neu20.mean, na.rm = TRUE))

p = ggplot(data = genes_avg)
for (celltype in c("iPS37", "NPC37", "Neu37")) {
  p = p + geom_point(aes_string(x = "length.cds",
                                y = paste("RE_", celltype, sep = ""),
                                colour = shQuote(celltype)),
                     alpha = 0.8, size = 2.4)
  p = p + geom_smooth(aes_string(x = "length.cds",
                                 y = paste("RE_", celltype, sep = ""),
                                 colour = shQuote(celltype)), 
                      se = FALSE, alpha = 0.5)
}
p = p + 
    ggthemes::theme_few() + 
    xlim(0, 10000) +
    ylim(0.5, 1.5) +
    xlab("CDS length") +
    ylab("Moving average RE")
p
```

RE dependence on CDS length. Analysis done on a gene level
```{r}
genes_avg = genes_stats %>%
            group_by(gene) %>%
            filter(length.cds == max(length.cds)) %>%
            mutate(length = floor(length.cds / 500)) %>%
            group_by(length) %>%
            summarise(length.cds = mean(length.cds, na.rm = TRUE), 
                      RE_iPS37 = mean(RE_iPS37.mean, na.rm = TRUE),
                      RE_NPC37 = mean(RE_NPC37.mean, na.rm = TRUE),
                      RE_Neu37 = mean(RE_Neu37.mean, na.rm = TRUE),
                      RE_NPC20 = mean(RE_NPC20.mean, na.rm = TRUE),
                      RE_Neu20 = mean(RE_Neu20.mean, na.rm = TRUE))

p = ggplot(data = genes_avg)
for (celltype in c("iPS37", "NPC37", "Neu37")) {
  p = p + geom_point(aes_string(x = "length.cds",
                                y = paste("RE_", celltype, sep = ""),
                                colour = shQuote(celltype)),
                     alpha = 1., size = 2.4)
  p = p + geom_smooth(aes_string(x = "length.cds",
                                 y = paste("RE_", celltype, sep = ""),
                                 colour = shQuote(celltype)), 
                      se = FALSE, alpha = 0.5)
}
p = p + 
    ggthemes::theme_few() + 
    xlim(0, 10000) +
    ylim(0, 3) +
    xlab("CDS length") +
    ylab("Moving average RE")
p
```

RE dependence on number of uAUGs.
```{r}
genes_avg = genes_stats %>%
            mutate(uatg = floor(uatg_count / 1)) %>%
            filter(uatg_count < 10) %>%
            group_by(uatg) %>%
            summarise(uatg_count = mean(uatg_count, na.rm = TRUE), 
                      RE_iPS37 = mean(RE_iPS37.mean, na.rm = TRUE),
                      RE_NPC37 = mean(RE_NPC37.mean, na.rm = TRUE),
                      RE_Neu37 = mean(RE_Neu37.mean, na.rm = TRUE),
                      RE_NPC20 = mean(RE_NPC20.mean, na.rm = TRUE),
                      RE_Neu20 = mean(RE_Neu20.mean, na.rm = TRUE))

p = ggplot(data = genes_avg)
for (celltype in c("iPS37", "NPC37", "Neu37")) {
  p = p + geom_point(aes_string(x = "uatg_count",
                                y = paste("RE_", celltype, sep = ""),
                                colour = shQuote(celltype)),
                     alpha = 1., size = 2.4)
  p = p + geom_smooth(aes_string(x = "uatg_count",
                                 y = paste("RE_", celltype, sep = ""),
                                 colour = shQuote(celltype)), 
                      se = FALSE, alpha = 0.5, method = "lm")
}
p = p + 
    ggthemes::theme_base() + 
    xlim(0, 10) +
    ylim(0.9, 1.1) +
    xlab("uAUG count") +
    ylab("Moving average RE")
p

summary(lm(RE_iPS37 ~ uatg_count, data = genes_avg))
summary(lm(RE_NPC37 ~ uatg_count, data = genes_avg))
summary(lm(RE_Neu37 ~ uatg_count, data = genes_avg))
```

RE dependence on number of uAUGs. Visualized with boxplots
```{r}
reshape2::melt(as.data.frame(genes_stats) %>%
               filter(uatg_count < 20) %>%
               mutate(uatg_marker = floor(uatg_count / 10)) %>%
               select(uatg_marker, contains("RE")) %>%
               select(uatg_marker, contains("mean")), 
               id.vars = "uatg_marker") %>%
  mutate(celltype = gsub("RE_|[.]mean", "", variable)) %>%
  ggplot(aes(x = uatg_marker, y = value, group = uatg_marker, colour = celltype, fill = celltype)) +
  geom_boxplot(alpha = 0.5) +
  ggthemes::theme_few() +
  scale_x_continuous(breaks = c(0, 1), labels = c("0-9", "10-20")) +
  coord_cartesian(ylim = c(0, 2)) +
  xlab("uAUG count") +
  ylab("Ribosomal engagement") +
  facet_wrap(~celltype, scales = "free")

```

RE dependence on GC percentage of a CDS. Analysis done on a transcript level
```{r}
genes_avg = genes_stats %>%
            mutate(gc = floor(gc_percentage.cds / 0.01)) %>%
            group_by(gc) %>%
            summarise(gc_percentage = mean(gc_percentage.cds, na.rm = TRUE), 
                      RE_iPS37 = mean(RE_iPS37.mean, na.rm = TRUE),
                      RE_NPC37 = mean(RE_NPC37.mean, na.rm = TRUE),
                      RE_Neu37 = mean(RE_Neu37.mean, na.rm = TRUE),
                      RE_NPC20 = mean(RE_NPC20.mean, na.rm = TRUE),
                      RE_Neu20 = mean(RE_Neu20.mean, na.rm = TRUE))

p = ggplot(data = genes_avg)
for (celltype in c("iPS37", "NPC37", "Neu37")) {
  p = p + geom_point(aes_string(x = "gc_percentage",
                                y = paste("RE_", celltype, sep = ""),
                                colour = shQuote(celltype)),
                     alpha = 0.8, size = 2.4)
  p = p + geom_smooth(aes_string(x = "gc_percentage",
                                 y = paste("RE_", celltype, sep = ""),
                                 colour = shQuote(celltype)), 
                      se = FALSE, alpha = 0.5)
}
p = p + 
    ggthemes::theme_few() + 
    xlim(0.3, 0.8) +
    ylim(0.5, 1.5) +
    xlab("GC percentage") +
    ylab("Moving average RE")
p
```

RE dependence on GC percentage of a CDS. Analysis done on a gene level
```{r}
genes_avg = genes_stats %>%
            group_by(gene) %>%
            filter(length.cds == max(length.cds)) %>%
            mutate(gc = floor(gc_percentage.cds / 0.01)) %>%
            group_by(gc) %>%
            summarise(gc_percentage = mean(gc_percentage.cds, na.rm = TRUE), 
                      RE_iPS37 = mean(RE_iPS37.mean, na.rm = TRUE),
                      RE_NPC37 = mean(RE_NPC37.mean, na.rm = TRUE),
                      RE_Neu37 = mean(RE_Neu37.mean, na.rm = TRUE),
                      RE_NPC20 = mean(RE_NPC20.mean, na.rm = TRUE),
                      RE_Neu20 = mean(RE_Neu20.mean, na.rm = TRUE))

p = ggplot(data = genes_avg)
for (celltype in c("iPS37", "NPC37", "Neu37")) {
  p = p + geom_point(aes_string(x = "gc_percentage",
                                y = paste("RE_", celltype, sep = ""),
                                colour = shQuote(celltype)),
                     alpha = 1., size = 2.4)
  p = p + geom_smooth(aes_string(x = "gc_percentage",
                                 y = paste("RE_", celltype, sep = ""),
                                 colour = shQuote(celltype)), 
                      se = FALSE, alpha = 0.5)
}
p = p + 
    ggthemes::theme_few() + 
    xlim(0.3, 0.8) +
    ylim(0, 3) +
    xlab("GC percentage") +
    ylab("Moving average RE")
p
```

RE dependence on 3'UTR length from QAPA estimates
```{r}
utrs_avg = utrs_tbl %>%
           mutate(length = floor(Length / 250)) %>%
           group_by(length) %>%
           summarise(length.utr = mean(Length, na.rm = TRUE), 
                     RE_iPS37 = mean(RE_iPS37.mean, na.rm = TRUE),
                     RE_NPC37 = mean(RE_NPC37.mean, na.rm = TRUE),
                     RE_Neu37 = mean(RE_Neu37.mean, na.rm = TRUE),
                     RE_NPC20 = mean(RE_NPC20.mean, na.rm = TRUE),
                     RE_Neu20 = mean(RE_Neu20.mean, na.rm = TRUE))

p = ggplot(data = utrs_avg)
for (celltype in c("iPS37", "NPC37", "Neu37")) {
  p = p + geom_point(aes_string(x = "length.utr",
                                y = paste("RE_", celltype, sep = ""),
                                colour = shQuote(celltype)),
                     alpha = 0.8, size = 2.4)
  p = p + geom_smooth(aes_string(x = "length.utr",
                                 y = paste("RE_", celltype, sep = ""),
                                 colour = shQuote(celltype)), 
                      se = FALSE, alpha = 0.5)
}
p = p + 
    ggthemes::theme_few() + 
    xlim(0, 10000) +
    xlab("3'UTR length") +
    ylab("Moving average RE") +
    scale_y_continuous(breaks = c(0.5, 1.0, 1.5), limits = c(0.5, 1.5))

p = p + 
    ggthemes::theme_few() + 
    xlab("3'UTR length") +
    ylab("Moving average RE") +
    scale_x_log10(limits = c(1000, 10000)) +
    scale_y_continuous(breaks = c(0.5, 1.0, 1.5), limits = c(0.5, 1.5))
p
```

