---
title: "Train a random forest classifier for RE predictions"
output:
  html_notebook:
    toc: true
    number_sections: true
    theme: readable
author: Marat Mufteev
date: "`r format(Sys.time(), 'Generated on %d %B, %Y')`"
---

Upload libraries
```{r}
library(dplyr)
library(caret)
library(parallel)
library(doParallel)
```

Current directory to generate data paths
```{r}
current_dir = getwd()
```

Upload RE and conserved binding elements data
```{r}
RE_path = paste(current_dir, 
                "/data/RE.3UTR.deseq.horizontal.quant_normalized.stats.tab", 
                sep = "")
UTR_fasta_path = paste(current_dir, 
                       "/data/qapa_hg19_3utrs.fa", 
                       sep = "")
elements_path = paste(current_dir, 
                      "/data/counts_design.ATTRACT.full.cons_0.9.csv", 
                      sep = "")

# Data on ATtRACT motifs and found binding elements
elements = read.csv(elements_path, header = T) %>%
           dplyr::rename(Name = X)

seqs = Biostrings::readDNAStringSet(UTR_fasta_path)
seq_features = data.frame(seqs@ranges)[,c("names", "width")] %>%
               dplyr::rename(Name = names,
                             L = width)

RE_seq = read.table(RE_path, header = T) %>%
         dplyr::rename(Name = UTR3_isoform) %>%
         mutate(RE_iPS37.mean = ifelse(RE_iPS37.noise < 1, RE_iPS37.mean, NA),
                RE_NPC37.mean = ifelse(RE_NPC37.noise < 1, RE_NPC37.mean, NA),
                RE_Neu37.mean = ifelse(RE_Neu37.noise < 1, RE_Neu37.mean, NA),
                RE_NPC20.mean = ifelse(RE_NPC20.noise < 1, RE_NPC20.mean, NA),
                RE_Neu20.mean = ifelse(RE_Neu20.noise < 1, RE_Neu20.mean, NA)) %>%
         right_join(seq_features, by = "Name") %>%
         right_join(elements, by = "Name") %>%
         mutate(RE_iPS37.loess = predict(loess(formula = RE_iPS37.mean ~ L, data = .), .$L),
                RE_NPC37.loess = predict(loess(formula = RE_NPC37.mean ~ L, data = .), .$L),
                RE_Neu37.loess = predict(loess(formula = RE_Neu37.mean ~ L, data = .), .$L),
                RE_NPC20.loess = predict(loess(formula = RE_NPC20.mean ~ L, data = .), .$L),
                RE_Neu20.loess = predict(loess(formula = RE_Neu20.mean ~ L, data = .), .$L)) %>%
         mutate(RE_iPS37.deviation = RE_iPS37.mean - RE_iPS37.loess,
                RE_NPC37.deviation = RE_NPC37.mean - RE_NPC37.loess,
                RE_Neu37.deviation = RE_Neu37.mean - RE_Neu37.loess,
                RE_NPC20.deviation = RE_NPC20.mean - RE_NPC20.loess,
                RE_Neu20.deviation = RE_Neu20.mean - RE_Neu20.loess)
```

Clean RE data. Analysis was done for multiple 3`UTR length bins to remove the 
effect of 3'UTR length on RE. Code below is done for an example 3'UTR length bin.
```{r}
L_bin = 200
cell = "iPS37"
input_tbl = RE_seq %>% 
              select(L, paste("RE_", cell, ".mean", sep = ""),
                     contains("ENSG")) %>%
              dplyr::rename(RE = paste("RE_", cell, ".mean", sep = "")) %>%
              filter(L > L_bin & L < (L_bin + 200)) %>%
              select(-L) %>%
              filter(!is.na(RE))
# If binding element data has NA substitute with 0.
input_tbl[is.na(input_tbl)] = 0
clean_tbl = data.frame(scale(input_tbl))
```

Define low and high RE classes and clean table
```{r}
class_tbl = clean_tbl %>%
                select(RE, contains("ENSG")) %>%
                mutate(top_Q = quantile(RE, 0.85),
                       bottom_Q = quantile(RE, 0.15)) %>%
                mutate(RE_delta = replace(RE, RE < top_Q & RE > bottom_Q, 0),
                       RE_delta = replace(RE_delta, RE <= bottom_Q, dplyr::first(bottom_Q)),
                       RE_delta = replace(RE_delta, RE >= top_Q, dplyr::first(top_Q))) %>%
                select(-RE) %>%
                mutate(RE_delta = as.character(RE_delta)) %>%
                mutate(RE_delta = replace(RE_delta, RE_delta == as.character(bottom_Q), "low"),
                       RE_delta = replace(RE_delta, RE_delta == "0", "neutral"),
                       RE_delta = replace(RE_delta, RE_delta == as.character(top_Q), "high")) %>% 
                filter(RE_delta %in% c("low", "high")) %>%
                select(RE_delta, contains("ENSG"))
NA_filter = c(TRUE, !is.na(colSums(class_tbl[,-c(1)])))
class_tbl = class_tbl[,NA_filter]
```

Split into train (70%) and test (30%) data
```{r}
indices = sample(x = 1:dim(class_tbl)[1], 
                 size = as.integer(0.7*dim(class_tbl)[1]),
                 replace = F)
train_tbl = class_tbl[indices, ]
test_tbl = class_tbl[-indices, ]
```

Initiate random forest training. CV to control for overfitting.
```{r}
number_of_cores = 2
cluster <- makeCluster(number_of_cores)
registerDoParallel(cluster)
train_control <- trainControl(method="cv", number=5, 
                              allowParallel = T)
class_model <- train(RE_delta ~ .,
                     data = train_tbl, 
                     trControl = train_control, 
                     method = "rf", 
                     importance = TRUE)
stopCluster(cluster)
registerDoSEQ()
```

Save trained model and predictions for further analysis.
```{r}
predictions = predict(class_model, test_tbl)
CM = confusionMatrix(data = factor(test_tbl$RE_delta), 
                     reference = predictions)
class_model$testData = test_tbl
class_model$pred = predict(class_model, test_tbl)
class_model$testAccuracy = CM$overall[1]
```
